---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# foodbank

<!-- badges: start -->
<!-- badges: end -->

foodbank provides the [USDA FoodData Central](https://fdc.nal.usda.gov/) Foundation Foods dataset as a set of data frames ready for analysis in R. It includes 436 foods with nutrient values, serving size portions, and food categories.

## Installation

You can install the development version of foodbank from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("hadley/foodbank")
```

## Datasets

foodbank provides six tables centered around `food`:

- `food`: 436 foundation foods with descriptions and categories.
  - Each food belongs to one `food_category` (join on `food_category_id = id`).
  - Each food has many rows in `food_nutrient` (join on `fdc_id`).
  - Some foods have rows in `food_portion` (join on `fdc_id`).
- `food_nutrient`: 19,828 nutrient values (per 100g of food). Each row links to one `nutrient` (join on `nutrient_id = id`) and one `food` (join on `fdc_id`).
- `nutrient`: 477 nutrient definitions with names and units.
- `food_portion`: 187 serving size measures with gram weights. Each row links to one `measure_unit` (join on `measure_unit_id = id`).
- `measure_unit`: 123 units used in portion measures.
- `food_category`: 28 food groups.

## Example

```{r example, message = FALSE}
library(foodbank)
library(dplyr)
```

Look up the nutrients in hummus:

```{r}
food |>
  filter(description == "Hummus, commercial") |>
  left_join(food_nutrient, by = "fdc_id") |>
  left_join(nutrient, by = c("nutrient_id" = "id")) |>
  select(description, name, amount, unit_name) |>
  head(10)
```

Find the foods highest in protein:

```{r}
nutrient |>
  filter(name == "Protein") |>
  left_join(food_nutrient, by = c("id" = "nutrient_id")) |>
  left_join(food, by = "fdc_id") |>
  select(description, amount, unit_name) |>
  arrange(desc(amount)) |>
  head(10)
```

## DuckDB

You can also create a DuckDB database backed by the parquet files, which is useful for SQL queries or working with the data via dbplyr:

```{r, eval = requireNamespace("duckdb", quietly = TRUE)}
con <- DBI::dbConnect(duckdb::duckdb())

parquet_dir <- system.file("parquet", package = "foodbank")
parquet_files <- list.files(parquet_dir, pattern = "\\.parquet$", full.names = TRUE)
for (f in parquet_files) {
  table_name <- tools::file_path_sans_ext(basename(f))
  DBI::dbExecute(con, sprintf("CREATE VIEW %s AS SELECT * FROM '%s'", table_name, f))
}

DBI::dbGetQuery(con, "
  SELECT f.description, n.name, fn.amount, n.unit_name
  FROM food f
  JOIN food_nutrient fn ON f.fdc_id = fn.fdc_id
  JOIN nutrient n ON fn.nutrient_id = n.id
  WHERE f.description = 'Hummus, commercial'
  ORDER BY fn.amount DESC
  LIMIT 8
")

DBI::dbDisconnect(con)
```

## Data source

The data comes from the [USDA FoodData Central](https://fdc.nal.usda.gov/) Foundation Foods dataset (December 2025 release). Foundation foods have nutrient values derived primarily by chemical analysis, with extensive underlying metadata.
